---
title: "Regression - Predicting Diamond Prices"
output: html_notebook
---

## Project Overview
A diamond distributor has recently decided to exit the market and has put up a set of 3,000 diamonds up for auction. Seeing this as a great opportunity to expand its inventory, a jewelry company has shown interest in making a bid. To decide how much to bid, I will use a large database of diamond prices to build a model to predict the price of a diamond based on its attributes. Then I will use the results of that model to make a recommendation for how much the company should bid.

## Understand The Data
There are two datasets. `diamonds.csv` contains the data used to build the regression model. `new_diamonds_new.csv` contains the data for the diamonds the company would like to purchase. Both datasets contain carat, cut, and clarity data for each diamond. Only the `diamonds.csv` dataset has prices. I'll be predicting prices for the `new_diamonds.csv` dataset.

Variables:
`Carat` represents the weight of the diamond, and is a numerical variable.
`Cut` represents the quality of the cut of the diamond, and falls into 5 categories: fair, good, very good, ideal, and premium. In project zero, these categories were represented by an ordinal variable, 1-5. You can decide to use the ordinal or categorical variable.
`Clarity` represents the internal purity of the diamond, and falls into 8 categories: I1, SI2, SI1, VS2, VS1, VVS2, VVS1, and IF (in order from least to most pure). In project zero, these categories were represented by an ordinal variable, 1-8. You can decide to use the ordinal or categorical variable.
`Color` represents the color of the diamond, and is rated D through J, with D being the most colorless (and valuable) and J being the most yellow.

## Load The Data
```{r warning=FALSE}
# Load packages
library(ggplot2)
library(tidyverse)
library(corrplot)
library(plotly)
library(caTools)
library(caret)
```

```{r}
# Load the data
diamonds <- read.csv("diamonds.csv")
new_diamonds <- read.csv("new-diamonds.csv")

# check data dimension
dim(diamonds)
dim(new_diamonds)
```

```{r}
# inspect first few rows
head(diamonds)
```

```{r}
# have a glimpse on data
glimpse(diamonds)
```

```{r}
# check data summary
summary(diamonds)
```

```{r}
summary(new_diamonds)
```


discuss data

## Data Wrangling
```{r}
# remove first column
diamonds <- diamonds[-1]
```

```{r}
# check the levels of color variable
levels(diamonds$color)

# create a new column for color ordinary data
diamonds$color_ord <- diamonds$color

# reverse levels of factor so that D is highest
diamonds$color_ord <- fct_rev(factor(diamonds$color_ord))

# change the value from categorical to numeric
diamonds$color_ord <- as.integer(diamonds$color_ord)
```

```{r}
# select numeric variables
diamonds <- select(diamonds, carat, cut, color, clarity, price)
```


```{r}
# inspect first few rows after data cleaning
head(diamonds)
```

## Build Regression Model
```{r}
# split data into train and test sets
set.seed(227)
sample <- sample.int(n = nrow(diamonds), size = floor(0.9*nrow(diamonds)), replace = F)
train <- diamonds[sample,]
test <- diamonds[-sample,]
```

```{r}
diamond_cor <- cor(diamonds)
```

```{r}
corrplot(diamond_cor, method = "pie", type = "upper")
```

### Univariate Model
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point() + geom_smooth(method = "lm")
```

```{r}
ggplot(diamonds, aes(x=cut_ord, y=price)) + geom_point() + geom_smooth(method = "lm")
```

```{r}
linearMod <- train(price ~ ., data = diamonds, method = 'lm', 
                   trControl = trainControl(
                     method = 'cv',
                     number = 10,
                     verboseIter = TRUE
                   ))
print(linearMod)
```

```{r}
summary(linearMod)
```

```{r}
# use the model to predict for test data
pricePred <- predict(linearMod, data=diamonds)
```

```{r}
mean(pricePred)
```


```{r}
# combine actual and predict data
actual_pred <- data.frame(cbind(actuals=diamonds$price, preds=pricePred))

# check correlation of actual and predicted data
cor(actual_pred)
```
```{r}
head(actual_pred, 20)
```

```{r}
summary(actual_pred)
```

```{r}
new_diamonds$price <- predict(linearMod, data=new_diamonds[-1])
```

